---
title: "CBB 752 - Project 1.3"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(knitr)
library(ggplot2)
library(peplib) # load blosum matrix
library(gplots)
```

```{r, include=FALSE}
x <- read.delim("Struct_annotation.txt", header = FALSE, as.is = TRUE,
                col.names=c("ENSG", "loc", "GERP",
                            "syn", "ENST", "change", "wildtype", "mutant"))
```

# Extract mutations
```{r}
getMutations <- function(str) {
  (str <- sub(";.*", "", str)); (str <- sub(",.*", "", str))
  (str <- sub("\\d+_\\d+_\\d+_", "", str))
  (from <- sub("->.", "", str));
  (to <- sub(".->", "", str))
  data.frame(from=from, to=to)
}

mutations <- getMutations(x[,"change"])
mutations$syn <- x$syn

table(mutations$from)
table(mutations$to)
```

# Charge Change

We score each change in the following manner:
```{r}
charge.chage.table = matrix(c(0,1,1,1,0,2,1,2,0), nrow = 3, ncol=3, byrow = TRUE)
row.names(charge.chage.table) <- c("Neutral", "Hydrophilic", "Hydrophobic" )
colnames(charge.chage.table) <- c("Neutral", "Hydrophilic", "Hydrophobic" )
kable(charge.chage.table, caption = "Amino Acid Hydrophobicity Change Table") 
```

Let's calculate the score according to this scoring system
```{r}
# There are many scales for AA hydrophobiciy. They following was adapted from:
# http://www.imgt.org/IMGTeducation/Aide-memoire/_UK/aminoacids/IMGTclasses.html
aa.hydrophobic <- c("A", "C", "I", "L", "M", "F", "W", "V")
aa.hydrophilic <- c("R", "N", "D", "Q", "E", "K")
aa.neutral     <- c("G", "H", "P", "S", "T", "Y")
charge_change <- function(values) {
  from <- values[1];  to <- values[2]; cc <- NA
  if(from %in% aa.neutral) {
    if(to %in% aa.neutral) { cc <- charge.chage.table["Neutral", "Neutral"] }
    if(to %in% aa.hydrophilic) { cc <- charge.chage.table["Neutral", "Hydrophilic"] }
    if(to %in% aa.hydrophobic) { cc <- charge.chage.table["Neutral", "Hydrophobic"] }
  }
  if(from %in% aa.hydrophilic) {
    if(to %in% aa.hydrophilic) { cc <- charge.chage.table["Hydrophilic", "Hydrophilic"] }
    if(to %in% aa.neutral) { cc <- charge.chage.table["Hydrophilic", "Neutral"] }
    if(to %in% aa.hydrophobic) { cc <- charge.chage.table["Hydrophilic", "Hydrophobic"] }
  }
  if(from %in% aa.hydrophobic) {
    if(to %in% aa.hydrophobic) { cc <- charge.chage.table["Hydrophobic", "Hydrophobic"] }
    if(to %in% aa.neutral) { cc <- charge.chage.table["Hydrophobic", "Neutral"] }
    if(to %in% aa.hydrophilic) { cc <- charge.chage.table["Hydrophobic", "Hydrophilic"] }
  }
  return(cc)
}

mutations$cc <- apply(mutations, 1, charge_change)

```

Plot of AA chagne in charge
```{r, message=FALSE, warning=FALSE}
qplot(mutations$cc,
      geom="histogram",
      main="Charge Change Histogram", xlab="",
      border="blue", fill="red")

```

# Size Change

As as estimate for size, we used the average volume of buried residues, calculated from the surface area of the side chain (Richards 1977; Baumann et al. 1989)
Adapted from:
http://www.proteinsandproteomics.org/content/free/tables_1/table08.pdf
```{r, message=FALSE, warning=FALSE}
aa.sizes <- list("A"=92 , "C"=106, "I"=169, "L"=168, "M"=171, "F"=203,
                 "W"=240, "V"=142, "R"=255, "N"=135, "D"=125, "Q"=161, 
                 "E"=155, "K"=171, "G"=66 , "H"=167, "P"=129, "S"=99 ,
                 "T"=122, "Y"=203, "*"=0)
mutations$size <- abs(as.numeric(unlist(aa.sizes[mutations$from])) -
                      as.numeric(unlist(aa.sizes[mutations$to])))
qplot(mutations$size,
      geom="histogram",
      main="Size diffence Histogram", xlab="",
      border="blue", fill="red")
```


# Blosum 90
```{r, message=FALSE, warning=FALSE}
data("blosum90")
blosum <- function(values) {
  b <- NA
  from <- values[1];  to <- values[2];
  if( from != "*" & to != "*") { b <- blosum90[from,to] }
  return(b)
}

mutations$blosum <- apply(mutations, 1, blosum)
qplot(mutations$blosum,
      geom="histogram",
      main="Blosum 90 Histogram", xlab="",
      border="blue", fill="red")
```

# GERP Score

```{r}
gerp <- function(str) {
  tmp <- sub("GERP=", "", str)
  tmp <- sub(";.*", "", tmp)
  return(as.numeric(tmp))
}

mutations$gerp <- gerp(x$GERP)

hist(mutations$gerp, probability = TRUE, las=3,
     main="GERP Score Histogram", xlab="",
     border="blue", col="red")

```

# Premature Stop
We have sevral premature stop cases. Let's see how short is the read compared to the reference.

```{r}
pre.l <- which(mutations$syn == "prematureStop")
pre <- x[pre.l, c("mutant", "wildtype")]
pre$missing.len <- as.numeric(nchar(pre$wildtype) - regexpr("\\*", pre$mutant))
```
We can see that we have 3 missing nonsense mutations
One with length of 15, and two other with lenght of 101
We can assume taht the longer the deletion is, the more deleterious it is.

# Summary Statistics

Let's see what is the frequency of each mutation
```{r}
mutations.l <- array()
for(i in 1:nrow(mutations)) {
  m <- paste(mutations$from[i],mutations$to[i], sep="")
  mutations.l <- c(mutations.l, m)
}

sort(table(mutations.l))

```

```{r}
mutations.mat <- matrix(0, nrow = length(unique(mutations$from)),
                        ncol = length(unique(mutations$to)))
rownames(mutations.mat) = sort(unique(mutations$from))
colnames(mutations.mat) = sort(unique(mutations$to), decreasing = FALSE)

for(i in 1:nrow(mutations)) {
  mutations.mat[mutations$from[i], mutations$to[i]] <-
    1 + mutations.mat[mutations$from[i], mutations$to[i]]
}

mutations.mat
heatmap.2(mutations.mat, col=colorRampPalette(c("blue", "red")), 
           density.info="none", trace="none", dendrogram="none", 
            symm=F,symkey=T,symbreaks=T, scale="none") 

```


# Deleterious Score (DScore)
We define Dscore as following (for discussion see the writing section)
$$DScore = \alpha Charge + \beta Size + \gamma GERP + \delta Blosum$$
Each one of the coeffiecients can be determined by the user.
We chose for example:
$$\alpha = \beta = \gamma = \delta = 1$$

```{r}
coeff <- list(c=1, s=1, b=1, g=1) # This we can get as user input
mutations$score <- coeff$c*mutations$cc + coeff$s*mutations$size + 
  coeff$b*mutations$blosum + coeff$g*mutations$gerp
m <- which.max(mutations$score)
print(paste("The most deleterious mutations is:", x$loc[m]))
```